\chapter{Acknowledgements}
\fancyhead[CO,CE]{\emph{Acknowledgements}}
I dedicate this book to Mary Adams-Sioss. She was the only person that really believed in me, and therefore helped me believe in myself, at a time when I needed it most. She taught me math, but more importantly, she taught me the joys of learning, which has led me to college and a lifetime of exploring my every curiosity and fascination. I never could have retired at 32 without all of this. It's amazing what one teacher can do for a student, just by treating him with the respect that every human being deserves. Thank you, Mrs. Adams.

I also thank the following:

Joe Dominguez and Vicki Robin, authors of \emph{Your Money or Your Life.} That book planted the seed that blossomed into my quest for financial independence. It persuaded me that it's possible, gave me a more positive connotation of frugality, and provided a road map to get me started on my journey. Ultimately, I found my own way, but their book was nonetheless foundational, and many of the ideas in this book are inspired by theirs.

Daniel Quinn, Ralph Waldo Emerson, Henry David Thoreau, and Alan Watts for writing books that profoundly shifted my vision and values, helped me work through my obsession with freedom, and addressed some of the most fundamental questions that vexed me. Their inspiration and influence can be felt in every page of this book.

My brother, for recommending \emph{Your Money or Your Life,} and for helping and supporting me in the early phases of this goal.

Stephanie Reyes, who believed in my retirement goal at a time when it was still quite fragile.

Cari Burstein, for her love, gentle encouragement, and ideas.

Of course, my thanks to all of my friends, who bring my life so much love and joy.
